Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-02-11 paper.Rnw
final 2-11-20?

@@

All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.

@@

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included \Sexpr{max(sapply(levels(m3),function(ll) sum(optmatch:::getZfromMatch(m3)[m3==ll])))} $H=1$ subjects matched to a
single $H=0$ subject, and another matched \Sexpr{max(sapply(levels(m3),function(ll) sum(!optmatch:::getZfromMatch(m3)[m3==ll])))} $H=0$ subjects to a
single $H=1$ subject.

@@

print(xtable(matchTEtab,
  align='rrllccc',
  caption='Estimates, standard errors, and sensitivity analysis of the weighted average effect of hint usage, under different weighting schemes.',
  label='tab:matchResults'),
  include.rownames=FALSE,sanitize.text.function =function(x) x,add.to.row=addtorow,include.colnames=FALSE)
@

>>>>

The \texttt{R} and Stan code used to for all of the analyses in the paper can be found in a github repository, \url{github.com/[redacted]}.

@@

All told, the analyses presented here were all based on \Sexpr{pn(NtFinal)} students assigned to the CTA1 condition; the estimation of direct effects in Section \ref{sec:mediation} and the principal stratification also relied on the \Sexpr{pn(Ntot-NtFinal)} students assigned to control.
Together, the \Sexpr{pn(Ntot)} students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
%The observational study (Section \ref{sec:observational} and part of the mediation analysis \ref{sec:mediation} used only the treatment (CTA1) group.

@@

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included \Sexpr{max(sapply(levels(m3),function(ll) sum(optmatch:::getZfromMatch(m3)[m3==ll])))} $H=1$ subjects matched to a
single $H=0$ subject, and another matched \Sexpr{max(sapply(levels(m3),function(ll) sum(!optmatch:::getZfromMatch(m3)[m3==ll])))} $H=0$ subjects to a
single $H=1$ subject.
All students in the CTA1 group of the dataset were included in the match.

@@

print(xtable(matchTEtab,
  align='rrllccc',
  caption='Estimates, standard errors, and sensitivity analysis of the weighted average effect of hint usage on posttest scores, under different weighting schemes.',
  label='tab:matchResults'),
  include.rownames=FALSE,sanitize.text.function =function(x) x,add.to.row=addtorow,include.colnames=FALSE)
@

### 2020-02-11 paper.tex
final 2-11-20?

All told, the main analysis included $n=$5,009 students, 2,091 of whom were assigned to the CTA1 condition and 2,918 of whom were assigned to control.
The students were nested within 116 teachers, in 43 schools across five states.

@@

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included 44 $H=1$ subjects matched to a
single $H=0$ subject, and another matched 117 $H=0$ subjects to a
single $H=1$ subject.

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Feb 10 10:49:52 2020
\begin{table}[ht]
\centering
\begin{tabular}{rllccc}
  \hline
  &&&&\multicolumn{2}{c}{Sensitivity Intervals}\\
 Weights&Estimate&Std. Error&CI&[Pretest]&[Ethnicity] \\
 \hline
OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.30,-0.05] \\ 
  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.29,0.04] \\ 
  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.31,0.03] \\ 
   \hline
\end{tabular}
\caption{Estimates and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

@@

\subsection{Stan Code}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## data{
## //Sample sizes
##  int<lower=1> N; // total number of obs (long format)
##  int<lower=1> nstud; // # students
##  int<lower=1> nsec; // # sections (problems nested in sections)
## 
## // indices
##  int<lower=1,upper=nstud> studentM[N]; //for each obs, which student?
##  int<lower=1,upper=nsec> section[N]; // for each obs, which section?
## 
## // data data
##  int<lower=0,upper=1> hint[N]; // for each obs, was a hint requested?
## 
## }
## parameters{
## 
##  vector[nstud] eta; // ability parameter
##  real<lower=0,upper=1> p1; // mixing proportion
##  //ordered[2] mu;             // locations of mixture components
##  real<upper=0> mu0; //location of mixture component 0
##  vector<lower=0>[2] sigma;  // scales of mixture components
## 
##  real delta[nsec]; // section difficulty parameter
##  //real<lower=0> sigSec; // scale of difficulty parameters
## 
## }
## transformed parameters{
##  real<lower=0> mu1; // location of mixture component 1
##  mu1= -(1-p1)/p1*mu0;
## }
## model{
##  // logs of mixing proportions to save time
##  real log_p1=log(p1);
##  real log_p0=log(1-p1);
## 
## // linear predictor for logit model
##  real linPred[N];
## 
## // priors
##  sigma ~ lognormal(0, 2);
##  mu0 ~ normal(0, 2);
## // sigSec ~ lognormal(0,2);
##  p1~ beta(2,2); // I'd rather the groups have similar sizes
## 
## // model for section difficulty parameters
##  delta~normal(0,3);//sigSec);
## 
## // Rasch model for hint requests
##  for(i in 1:N)
##   linPred[i]= eta[studentM[i]]-delta[section[i]];
## 
##  hint~bernoulli_logit(linPred);
## 
## 
## // mixture model for ability parameters
##  for(n in 1:nstud)
##   target += log_sum_exp(log_p0+normal_lpdf(eta[n]|mu0,sigma[1]),
## 			log_p1+normal_lpdf(eta[n]|mu1,sigma[2]));
## 
## }
\end{verbatim}
\end{kframe}
\end{knitrout}
%\clearpage
\section{Covariate Balance Before and After Matching}
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Feb 10 10:56:04 2020
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  &\multicolumn{3}{c}{Unmatched}&\multicolumn{3}{c}{Matched}\\
 \hline
 & std.diff & z &  & std.diff & z &  \\ 
  \hline
grade9 & -0.34 & -7.05 & *** & 0.03 & 0.44 &     \\ 
  gradehigher & 0.34 & 7.05 & *** & -0.03 & -0.44 &     \\ 
  raceWhiteAsian & -0.64 & -12.83 & *** & 0.03 & 0.75 &     \\ 
  raceBlackMulti & 0.23 & 4.71 & *** & 0.01 & 0.24 &     \\ 
  raceHispAIAN & 0.51 & 10.35 & *** & -0.05 & -0.98 &     \\ 
  xirt & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
  spectypical & -0.16 & -3.32 & *** & 0.03 & 0.48 &     \\ 
  specspeced & 0.34 & 7.07 & *** & -0.10 & -1.44 &     \\ 
  specgifted & -0.14 & -3.03 & **  & 0.06 & 1.16 &     \\ 
  esl0 & -0.37 & -7.70 & *** & 0.04 & 0.63 &     \\ 
  esl1 & 0.37 & 7.70 & *** & -0.04 & -0.63 &     \\ 
  stateCT & 0.58 & 11.81 & *** & -0.00 & 0.00 &     \\ 
  stateKY & -0.36 & -7.51 & *** & -0.00 & 0.00 &     \\ 
  stateLA & -0.08 & -1.68 & .   & -0.00 & 0.00 &     \\ 
  stateMI & -0.36 & -7.51 & *** & 0.00 & 0.00 &     \\ 
  stateTX & 0.43 & 8.77 & *** & 0.00 & 0.00 &     \\ 
   \hline
\end{tabular}
\caption{Covariate balance (standardized differences and Z-scores) before and after the propensity score match.} 
\label{tab:balance}
\end{table}

>>>>

All told, the analyses presented here were all based on 2,091 students assigned to the CTA1 condition; the estimation of direct effects in Section \ref{sec:mediation} and the principal stratification also relied on the 2,918 students assigned to control.
Together, the 5,009 students were nested within 116 teachers, in 43 schools across five states.
%The observational study (Section \ref{sec:observational} and part of the mediation analysis \ref{sec:mediation} used only the treatment (CTA1) group.

@@

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included 44 $H=1$ subjects matched to a
single $H=0$ subject, and another matched 117 $H=0$ subjects to a
single $H=1$ subject.
All students in the CTA1 group of the dataset were included in the match.

@@

% latex table generated in R 3.6.2 by xtable 1.8-4 package
% Tue Feb 11 09:12:59 2020
\begin{table}[ht]
\centering
\begin{tabular}{rllccc}
  \hline
  &&&&\multicolumn{2}{c}{Sensitivity Intervals}\\
 Weights&Estimate&Std. Error&CI&[Pretest]&[Ethnicity] \\
 \hline
OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.30,-0.05] \\ 
  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.29,0.04] \\ 
  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.31,0.03] \\ 
   \hline
\end{tabular}
\caption{Estimates, standard errors, and sensitivity analysis of the weighted average effect of hint usage on posttest scores, under different weighting schemes.} 
\label{tab:matchResults}
\end{table}

@@

\subsection{Stan Code}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## data{
## //Sample sizes
##  int<lower=1> N; // total number of obs (long format)
##  int<lower=1> nstud; // # students
##  int<lower=1> nsec; // # sections (problems nested in sections)
## 
## // indices
##  int<lower=1,upper=nstud> studentM[N]; //for each obs, which student?
##  int<lower=1,upper=nsec> section[N]; // for each obs, which section?
## 
## // data data
##  int<lower=0,upper=1> hint[N]; // for each obs, was a hint requested?
## 
## }
## parameters{
## 
##  vector[nstud] eta; // ability parameter
##  real<lower=0,upper=1> p1; // mixing proportion
##  //ordered[2] mu;             // locations of mixture components
##  real<upper=0> mu0; //location of mixture component 0
##  vector<lower=0>[2] sigma;  // scales of mixture components
## 
##  real delta[nsec]; // section difficulty parameter
##  //real<lower=0> sigSec; // scale of difficulty parameters
## 
## }
## transformed parameters{
##  real<lower=0> mu1; // location of mixture component 1
##  mu1= -(1-p1)/p1*mu0;
## }
## model{
##  // logs of mixing proportions to save time
##  real log_p1=log(p1);
##  real log_p0=log(1-p1);
## 
## // linear predictor for logit model
##  real linPred[N];
## 
## // priors
##  sigma ~ lognormal(0, 2);
##  mu0 ~ normal(0, 2);
## // sigSec ~ lognormal(0,2);
##  p1~ beta(2,2); // I'd rather the groups have similar sizes
## 
## // model for section difficulty parameters
##  delta~normal(0,3);//sigSec);
## 
## // Rasch model for hint requests
##  for(i in 1:N)
##   linPred[i]= eta[studentM[i]]-delta[section[i]];
## 
##  hint~bernoulli_logit(linPred);
## 
## 
## // mixture model for ability parameters
##  for(n in 1:nstud)
##   target += log_sum_exp(log_p0+normal_lpdf(eta[n]|mu0,sigma[1]),
## 			log_p1+normal_lpdf(eta[n]|mu1,sigma[2]));
## 
## }
\end{verbatim}
\end{kframe}
\end{knitrout}
%\clearpage
\section{Covariate Balance Before and After Matching}
% latex table generated in R 3.6.2 by xtable 1.8-4 package
% Tue Feb 11 09:12:59 2020
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  &\multicolumn{3}{c}{Unmatched}&\multicolumn{3}{c}{Matched}\\
 \hline
 & std.diff & z &  & std.diff & z &  \\ 
  \hline
grade9 & -0.34 & -7.05 & *** & 0.03 & 0.44 &     \\ 
  gradehigher & 0.34 & 7.05 & *** & -0.03 & -0.44 &     \\ 
  raceWhiteAsian & -0.64 & -12.83 & *** & 0.03 & 0.75 &     \\ 
  raceBlackMulti & 0.23 & 4.71 & *** & 0.01 & 0.24 &     \\ 
  raceHispAIAN & 0.51 & 10.35 & *** & -0.05 & -0.98 &     \\ 
  xirt & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
  spectypical & -0.16 & -3.32 & *** & 0.03 & 0.48 &     \\ 
  specspeced & 0.34 & 7.07 & *** & -0.10 & -1.44 &     \\ 
  specgifted & -0.14 & -3.03 & **  & 0.06 & 1.16 &     \\ 
  esl0 & -0.37 & -7.70 & *** & 0.04 & 0.63 &     \\ 
  esl1 & 0.37 & 7.70 & *** & -0.04 & -0.63 &     \\ 
  stateCT & 0.58 & 11.81 & *** & -0.00 & 0.00 &     \\ 
  stateKY & -0.36 & -7.51 & *** & -0.00 & 0.00 &     \\ 
  stateLA & -0.08 & -1.68 & .   & -0.00 & 0.00 &     \\ 
  stateMI & -0.36 & -7.51 & *** & 0.00 & 0.00 &     \\ 
  stateTX & 0.43 & 8.77 & *** & 0.00 & 0.00 &     \\ 
   \hline
\end{tabular}
\caption{Covariate balance (standardized differences and Z-scores) before and after the propensity score match.} 
\label{tab:balance}
\end{table}
